One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-11 paper/model.tex
intro draft completed
updating 3d write up

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine two-dimensional
(2D, or razor-thin) discs in favour of numerical 
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Thus, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

@@

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $R_0$ is a reference cylindrical radius and $s$ is the power-law
index describing the smooth disc, and $\Sigma_\mathrm{ref}$ is a
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
$R_{2}$. The bump function mimicks mass built-up at some location in 
the disc.

@@

\section{Numerical methods}
We adopt computational units such that $G=M_*=R_0=1$. Time is measured
in the Keplerian orbital period at the reference radius, $P_0\equiv 
2\pi/\Omega_k(R_0)$.

>>>>

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine two-dimensional
(2D, or razor-thin) discs in favour of numerical 
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Thus, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

@@

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $s$ is the power-law index describing the smooth disc and
$\Sigma_\mathrm{ref}$ is a 
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
$R_{2}$. The bump function mimicks mass built-up at some location in 
the disc.

@@

\section{Numerical methods}\label{methods}
We adopt computational units such that $G=M_*=R_0=1$. Time is measured
in the Keplerian orbital period at the reference radius, $P_0\equiv 
2\pi/\Omega_k(R_0)$.

### 2014-11-11 paper/results_3d.tex
updating 3d write up

\section{Three-dimensional simulations}\label{results3d}
In this section we briefly review 3D simulations carried 
out using ZEUS-MP and PLUTO. The main purpose is to check 
the above results with different numerical codes, and validate  
the 2D approximation.

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent 
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
$H$in radius. 
Because of the much reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
the 2D simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=100P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. We also show results from strictly
isothermal discs ($q=0$), which clearly display no growth. %why no high m
                                %modes as in 2D for iso disc?
                                %resolution?
Results from ZEUS-MP appears to be off-set from that obtained using
PLUTO. We suspect this may be due to larger numerical noise in the
ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
end of the run. However, we measure similar growth rates,
\begin{align*}
&\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
&\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations above. This may be due to the
low resolutions adopted. 
%growth rate does not depend on pert amp

@@

Visualisations of the 3D simluations are shown in
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the spiral 
amplitudes reached in the two codes are similar. Although the
non-axisymmetric density has similar amplitudes at both heights, a
coherent spiral is less evident away from the disc midplane. 
The agreement between the two codes, as well as with the 2D
simulations, is satisfactory.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_z0}
      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/polarxy_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.33]{figures/pdisk_020_z0}
      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/pdisk_020_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The non-axisymmetric density at the midplane (left) and
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane.\label{3d_prelim}}   
\end{figure}

\subsection{Vertical motion}
We measure three-dimensionality using the parameter $\Theta$ as
defined in Eq. \ref{theta}.  Vertical velocities
do increase with the growth of the one-arm spiral, 
%Larger amplitudes
%are reached in ZEUS because of larger numerical noise. 
but we find  $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
compared to horizontal motions. This supports the 2D approximation.

%\begin{figure}
%  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
% \caption{Evolution of three-dimensionality $\Theta$, as defined by
%    Eq. \ref{theta}, which corresponds to the
%   average vertical kinetic energy density, scaled by 
%    the average horizontal kinetic energy.\label{3d_vert}} 
%\end{figure}   
\subsection{Angular momentum evolution}

Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. ZEUS-MP does not conserve angular momentum very well, but
the variation $|\Delta J/J|< O(10^{-5})$ is not significant compared
to the individual components $|\Delta J_{0,1}/J|\sim 
6\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conervation, with $|\Delta
J/J|=O(10^{-8})$.

>>>>

\section{Three-dimensional simulations}\label{results3d}
In this section we review 3D simulations carried 
out using ZEUS-MP and PLUTO. The main purpose is to verify 
the above results with different numerical codes, and validate  
the 2D approximation.

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
$H$in radius. Because of the much reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
the 2D simulation.

The 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$ and damp
merdional velocities near the radial boundaries.

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. We also show results from strictly
isothermal discs ($q=0$), which clearly display no growth. %why no high m
                                %modes as in 2D for iso disc?
                                %resolution?
The mode amplitude from ZEUS-MP appears to be off-set from that obtained using
PLUTO. We suspect this may be due to larger numerical noise in the
ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
end of the run. However, we measure similar growth rates,
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations above. This may be due to the
low resolutions adopted and or the small softening length employed in
2D.  
%growth rate does not depend on pert amp

@@

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the spirals in the two codes
have reached similar amplitudes. %The agreement between the codes,
% as well as with the 2D simulations, is satisfactory.    
Away from the disc midplane, a coherent spiral arm is
not prominent, but the disturbance is radially more global, extending
up to $R\sim 4R_0$. %implications?

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.382]{figures/polarxy_dens015_z0}
      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/polarxy_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.382]{figures/pdisk_020_z0}
      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/pdisk_020_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The non-axisymmetric density at the midplane (left) and
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane.\label{3d_prelim}}   
\end{figure}

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-arm
spiral in the PLUTO run. The main spiral is confined to $z < H$.
However, outside $R>2R_0$ we observe spirals

\begin{figure}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm 0cm]{figures/pdisk_rz_020_sg.ps}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm 0.64cm]{figures/pdisk_rz_020_nsg.ps}
  \caption{Non-axisymmetric density in the meridional plane from the
    PLUTO simulation. The slices are taken at the azimuth of   
    $\mathrm{max}[\Delta\rho(r,\pi/2,\phi)]$. Arrows correspond to the vector 
    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
    to the inner (outer) portions of the disc. 
    \label{3d_rz}} 
\end{figure}

We measure three-dimensionality using the parameter $\Theta$ as
defined in Eq. \ref{theta}.  Vertical velocities
do increase with the growth of the one-arm spiral, 
%Larger amplitudes
%are reached in ZEUS because of larger numerical noise. 
but we find  $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
compared to horizontal motions. This supports the 2D approximation.

%\begin{figure}
%  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
% \caption{Evolution of three-dimensionality $\Theta$, as defined by
%    Eq. \ref{theta}, which corresponds to the
%   average vertical kinetic energy density, scaled by 
%    the average horizontal kinetic energy.\label{3d_vert}} 
%\end{figure}

\subsection{Angular momentum evolution}

Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. ZEUS-MP does not conserve angular momentum very well, but
the variation $|\Delta J/J|< O(10^{-6})$ is not significant compared
to the individual components $|\Delta J_{0,1}/J|\sim 
5\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conervation, with $|\Delta
J/J|=O(10^{-8})$.
